k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-02-29 paper.Rmd
abstract

    When assigning stimuli to experimental conditions, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) to obtain a stimulus partitioning that simultaneously achieves high values according to two criteria: the diversity, corresponding to between-set similarity, and the dispersion, corresponding to pairwise within-set dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We present a novel exact algorithm that returns (multiple) globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, it scales to rather large data sets and can be applied to solve most real-world use cases. In a simulation study, we evaluate several alternatives to combine our optimal approach with the BILS heuristic, which -- due to its bicriterion approach -- naturally lends itself to providing high diversity on top of an optimal dispersion. We also provide two further extensions to the BILS: First, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Second, we allow that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. The original BILS as well as our extensions are freely available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.

>>>>

    When assigning stimuli to experimental conditions, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion algorithm (BILS) to obtain stimulus partitionings that simultaneously satisfy two anticlustering criteria: the diversity, corresponding to between-set similarity, and the dispersion, corresponding to pairwise within-set dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We present a novel exact algorithm that returns (multiple) globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets and can be applied to solve most real-world use cases. In a simulation study, we evaluate several options to combine our optimal approach with the BILS heuristic, which -- due to its bicriterion approach -- naturally lends itself to providing high diversity on top of an optimal dispersion. We also provide two further extensions to the BILS: First, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Second, we allow that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. The original BILS as well as our extensions are freely available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
